A recipe for extremely reproducible enrichment analysis

### 2023-03-24 manuscript.Rmd
updates

---
title: "A recipe for extremely reproducible enrichment analysis"
author: "Mark Ziemann and Anusuiya Bora"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
  html_document:
    toc: true
    toc_float: true
    fig_width: 7
    fig_height: 5
theme: cosmo
---

<b>Alt titles:<b>

@@

```{r,libs}

@@

@@

```{r,diagram,fig.cap = "Figure XX: Flow chart of example workflow."}

@@

We use this analysis to try to understand better the effect of DNA methylatransferase inhibitors on
genome regulation and elaborate upon previous findings by Lund et al (2014), which mainly focused on
the relationship between RNA expression and DNA methylation, but did not show results of a
transcriptome-only enrichment analysis.

Furthermore, we provide a comparison of the two most common approaches for enrichment analysis; ORA and
FCS, and give guidance for interpreting results of such tests.

After fetching the data from DEE2, the number of reads per sample was confirmed to be in the range of
11-16M.
Genes with fewer than 10 reads per sample on average were removed from the analysis, leaving 13,168
genes.
Differential expression analysis revealed 3,598 differentially expressed genes with FDR<0.05, with
1,672 exhibiting higher expression and 1,926 exhibiting lower expression in 5-azacytidine treated cells.

These differentially expressed gene sets were subjected to ORA analysis separately, using the list of
13,168 genes as the background, and gene sets from REACTOME.

With ORA, we observed 149 upregulated and 48 downregulated gene sets.

We prioritised the enrichment analysis by enrichment score, a pseudo measure for effect size,
which tends to highlight smaller and more specific gene sets undergoing more drastic differential
expression, rather than large gene sets that undergo very small expression changes.
This led to the identification of upregulated pathways related to nuclear RNA export,
glucokinase regulation and G1/S DNA replication pathways; while downregulated pathways
were associated with interferon signaling, interleukin signaling, granulopoiesis and NADPH oxidase
(host defence).

After FCS, we identified 241 upregulated genesets and 155 downregulated genesets.
FCS upregulated pathways were associated with prometaphase chromosome condensation, nuclear RNA export
and defective homologous recombination repair.
Downregulated pathways related to endosomal/lysosome pathway, interferon signaling and TLR/TNF signaling.

Regarding upregulated pathways, the similarity between ORA and FCS was relatively high, exhibiting a
Jaccard index of 0.54, however these method disagreed in downregulated pathways, with a Jaccard
index of only 0.27.

@@

>>>>

---
title: "A recipe for extremely reproducible enrichment analysis"
author: "Mark Ziemann and Anusuiya Bora"
date: "`r Sys.Date()`"
output:
  rmdformats::robobook:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 5
---

<b>Alt titles:</b>

@@

```{r,libs,echo=FALSE}

@@

![Figure XX: The computational research reproducibility spectrum. Creator: Altuna Akalin.](img/repro.webp "The computational research reproducibility spectrum.")

@@

```{r,diagram,echo=FALSE,fig.cap = "Figure XX: Flow chart of example workflow."}

@@

We use this analysis to try to understand better the effect of DNA methylatransferase inhibitors on
genome regulation and elaborate upon previous findings by Lund et al (2014), which mainly focused on
the relationship between RNA expression and DNA methylation, but did not show results of a
transcriptome-only enrichment analysis.
Furthermore, we provide a comparison of the two most common approaches for enrichment analysis; ORA and
FCS, and give guidance for interpreting results of such tests.

After fetching the data from DEE2, the number of reads per sample was confirmed to be in the range of
11-16M.
Genes with fewer than 10 reads per sample on average were removed from the analysis, leaving 13,168
genes.
Differential expression analysis revealed 3,598 differentially expressed genes with FDR<0.05, with
1,672 exhibiting higher expression and 1,926 exhibiting lower expression in 5-azacytidine treated cells
(Figure XXA).
These differentially expressed gene sets were subjected to ORA analysis separately, using the list of
13,168 genes as the background, and gene sets from REACTOME.

With ORA, we observed 149 upregulated and 48 downregulated gene sets.
We prioritised the enrichment analysis by enrichment score, a pseudo measure for effect size,
which tends to highlight smaller and more specific gene sets undergoing more drastic differential
expression, rather than large gene sets that undergo very small expression changes.
This led to the identification of upregulated pathways related to nuclear RNA export,
glucokinase regulation and G1/S DNA replication pathways; while downregulated pathways
were associated with interferon signaling, interleukin signaling, granulopoiesis and NADPH oxidase
(host defence) (Figure XXB).

After FCS, we identified 241 upregulated genesets and 155 downregulated genesets.
FCS upregulated pathways were associated with prometaphase chromosome condensation, nuclear RNA export
and defective homologous recombination repair (Figure XXC).
Downregulated pathways related to endosomal/lysosome pathway, interferon signaling and TLR/TNF signaling.

Regarding upregulated pathways, the similarity between ORA and FCS was relatively high, exhibiting a
Jaccard index of 0.54, however these method disagreed in downregulated pathways, with a Jaccard
index of only 0.27 (Figure XXD).

![Figure XX: Differential expression and enrichment analysis with the example workflow.](img/fig_example.png "Example enrichment analysis on")

@@

## Supplementary information
